---
title: "NFL Predictor"
output: html_document
date: "2025-11-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, let's get the data
```{r}
library(nflfastR)
library(nflreadr)
library(dplyr)
library(tidyverse)
```

We will use the following features

**Note: Every AVG is a 5-game average**

Offense:

- AVG Rush Yards
- AVG Pass Yards
- injured QB1
- Third down conversion rate

Defense:

- AVG Opp yards allowed
- AVG Opp points allowed

Team Health

- IR Percentage
- Days of Rest 

Momentum/Extra

- Home Field Advantage
- SRS
- Point Differential (5 games)
- Turnover Differential (5 games)
- Wins in last 5 games

---

Optimized Version

```{r}
prepare_season_df <- function(pbp, season) {

  season_games <- unique(pbp$game_id[pbp$season == season])

  if (length(season_games) == 0) {
    stop(paste("No games found for season", season))
  }

  n <- length(season_games)

  # Pre-allocate
  game.id <- character(n)
  home.team <- away.team <- character(n)
  home.score <- away.score <- numeric(n)
  home.rush <- away.rush <- numeric(n)
  home.pass <- away.pass <- numeric(n)
  home.turnovers <- away.turnovers <- numeric(n)
  home.team.win <- integer(n)

  for (i in seq_len(n)) {

    gid <- season_games[i]
    dat <- pbp[pbp$game_id == gid, ]

    h <- dat$home_team[1]
    a <- dat$away_team[1]

    game.id[i] <- gid
    home.team[i] <- h
    away.team[i] <- a

    home.score[i] <- max(dat$total_home_score, na.rm = TRUE)
    away.score[i] <- max(dat$total_away_score, na.rm = TRUE)

    # Rush / Pass yards
    home.rush[i] <- sum(dat$yards_gained[dat$play_type=="run"  & dat$posteam==h], na.rm=TRUE)
    away.rush[i] <- sum(dat$yards_gained[dat$play_type=="run"  & dat$posteam==a], na.rm=TRUE)

    home.pass[i] <- sum(dat$yards_gained[dat$play_type=="pass" & dat$posteam==h], na.rm=TRUE)
    away.pass[i] <- sum(dat$yards_gained[dat$play_type=="pass" & dat$posteam==a], na.rm=TRUE)

    # ✅ Correct turnover definition
    home.turnovers[i] <- sum(dat$interception == 1 & dat$defteam == a, na.rm = TRUE) +
                          sum(dat$fumble_lost == 1 & dat$posteam == h, na.rm = TRUE)

    away.turnovers[i] <- sum(dat$interception == 1 & dat$defteam == h, na.rm = TRUE) +
                          sum(dat$fumble_lost == 1 & dat$posteam == a, na.rm = TRUE)

    home.team.win[i] <- as.integer(home.score[i] > away.score[i])
  }

  df <- data.frame(
    game.id,
    season = season,
    home.team,
    away.team,
    home.score,
    away.score,
    home.rush,
    away.rush,
    home.pass,
    away.pass,
    home.turnovers,
    away.turnovers,
    home.team.win,
    stringsAsFactors = FALSE
  )

  return(df)
}

```


Create Dataframes with nfl data
```{r}
pdb.train <- load_pbp(2000:2013)
pdb.2014 <- load_pbp(2014) # Done 
pdb.2015 <- load_pbp(2015) # Done
pdb.2016 <- load_pbp(2016) # Done
pdb.2017 <- load_pbp(2017) # Done
pdb.2018 <- load_pbp(2018) # Done
pdb.2019 <- load_pbp(2019) # Done 
pdb.2020 <- load_pbp(2020) # Done
pdb.2021 <- load_pbp(2021) # Done
pdb.2022 <- load_pbp(2022) # Done
pdb.2023 <- load_pbp(2023) # Done
pdb.2024 <- load_pbp(2024) # Done
```

Model Diagnostic
```{r}
# Build Training and Testing
#pbp.14.22 <- load_pbp(2022) # <-------------------------- Only Use when Changing Seasons
train.df <- prepare_season_df(pdb.train, 2010:2013)
test.df  <- prepare_season_df(pdb.2024, 2024) # <------ Change This for Each Year

# Fit Logistic Model
logit.model <- glm(
  home.team.win ~ home.turnovers + home.rush + home.pass +
                  away.turnovers + away.rush + away.pass,
  data = train.df,
  family = "binomial"
)

# Fit Random Forest
library(randomForest)

rf.model <- randomForest(
  as.factor(home.team.win) ~ home.turnovers + home.rush + home.pass +
                             away.turnovers + away.rush + away.pass,
  data = train.df,
  ntree = 500,
  mtry = 3
)


# Predict on 2023 Test Set

# LOGISTIC REGRESSION PREDICTIONS
test.df$logit_prob  <- predict(logit.model, newdata = test.df, type = "response")
test.df$logit_class <- as.integer(test.df$logit_prob > 0.5)



# RANDOM FOREST PREDICTIONS
rf.pred <- predict(rf.model, newdata = test.df, type = "prob")
test.df$rf_prob  <- rf.pred[, "1"]
test.df$rf_class <- as.integer(test.df$rf_prob > 0.5)

```

Prediction Dataframes
```{r}
# RF
rf.pred.2024 <- data.frame(
  game_id       = test.df$game.id,
  season        = 2024,
  model_version = "v1.1_rf_turnrushpass",
  pred_prob     = test.df$rf_prob,
  pred_class    = test.df$rf_class,
  true_class    = test.df$home.team.win,
  created_at    = Sys.time()
)

rf.pred.2024$congruence = ifelse(rf.pred.2024$pred_class == rf.pred.2024$true_class, 1, 0)

# Logit
logit.pred.2024 <- data.frame(
  game_id       = test.df$game.id,
  season        = 2024,
  model_version = "v1.1_logit_turnrushpass",
  pred_prob     = test.df$logit_prob,
  pred_class    = test.df$logit_class,
  true_class    = test.df$home.team.win,
  created_at    = Sys.time()
)

logit.pred.2024$congruence = ifelse(logit.pred.2024$pred_class == logit.pred.2024$true_class, 1, 0)

# Coin Flip
coin.flip.2024 <- data.frame(
  game_id       = test.df$game.id,
  season        = 2024,
  model_version = "v1.1_coin_turnrushpass",
  pred_class    = rbinom(nrow(test.df), 1, 0.5),
  true_class    = test.df$home.team.win,
  created_at    = Sys.time()
)

coin.flip.2024$congruence = ifelse(coin.flip.2024$pred_class == coin.flip.2024$true_class, 1, 0)
```

All Predictions
```{r}
all.pred.2024 <- bind_rows(
  rf.pred.2024,
  logit.pred.2024,
  coin.flip.2024
)
```


Save to Folder
```{r}
library(openxlsx)

write.xlsx(
  all.pred.2024,
  file = "predictions_2024.xlsx",
  overwrite = TRUE
)

```

